---
title: "Project Checkpoint 1"
author: "Marshall Graham"
date: "November 24, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#1
#Group members: Marshall, Zain, Colleen, Vishal

Main Questions:
1. What is the correlation between weather and ridership on Metro Transit and Niceride?
*How do bikeshare ridership patterns vary depending on temperature compared to buses?
*What other weather factors besides temperature influence ridership of Niceride and Metro Transit?

2. Is there a correlation between proximity of  a Niceride Station to a bus stop and usage patterns of Niceride stations?
*How do Niceride usage patterns vary over the course of one weekday? (Do the stations less than 500 meters from a bus stop have spikes in usage when a bus lets out?)

#2
#a. bikeshare patterns

#b. Research questions:
1. How do bikeshare usage patterns vary depending on weather compared to buses?
2. Is there a correlation between proximity of a Niceride station to a bus stop and usage patterns of the Niceride station?

#ALTERNATIVE QUESTIONS: 
-How do usage patterns of bikes and public transit vary from the norm on public holidays?
-Is metro transit usage affected when the bikeshare closes down for the year (or is it just seasonal)?

#Datasets we will use

Data Source: metaweather
Data Description: weather patterns for a certain day in Minneapolis
Data Limitations: challenging to get at all 365 days with ease (without going through them one at a time)
Data dimensions: 71x15 (for one day,)


#metroStops
Data source: Metro Transit
Data description: Location of each bus stop (with site ID, city, latitude, longitude, etc.)
Data limitations: bus stops that exist from 2014-2017
Data dimensions: 14,919 x 12

#metroRidership
Data source: Metro Transit
Data description: Gives the day, if it's a holiday, route number, route type, number of trips, and total number of riders on that day 
Data limitations: from January 2014- October 2017
Data dimensions: 131,078 x 10

#NicerideRidership
Data Source: Niceride 2016 data
Data Description: Data contains information about trips: the start and end stations of each trip, the total time it took to complete the journey, and whether the passenger was a casual rider or a member.
Data Limitations: Does not contain information about metro rides in December, January, February, and March. Only has 432283 rows with data in them, the rest are empty and this data has to be cleaned accordingly before use. 
Data dimensions: >432,284x8

#NiceRideStops
Data Source: Nice Ride 2016 data
Data Description: Data contains names of Metro Bus Stations, and the exact coordinates of each station (latitude and longitude)
Data Limitations: This dataset contains everything we require to come up with comprehensive visualizations hence, there are no limitations to this data. 
Data Dimensions: 202 x 6 


MinneapolisWeather
Data Source: "http://www.dnr.state.mn.us/climate/twin_cities/listings.html", the Department of Natural Resources has data going back to 1871 on the weather in Minneapolis/St. Paul. All data after 1938 is from MSP airport
Data Description: Gives the date, max temperature, min temperature, amount of percpititaion, amount of snow, and amount of snow on the ground
Data Limitations The specific data we're using is the weather data between 2010 and 2017. It's missing the average tempearture for a day, but the min and max are probably enough to work with
Data dimensions: 2884 x 6


```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(urltools)
library(jsonlite)
library(ggmap)
library(rvest)
library(geosphere)

```






Read in data on NiceRide, the bike share for the twin cities
```{r, cache=TRUE, message = FALSE, warning = FALSE}

niceRide2016<-read_csv("Nice_ride_trip_history_2016_season.csv")
#Throw out empty rows
niceRide2016<-niceRide2016%>%
   filter(!is.na(`Start date`))

niceRideLocations<-read_csv("Nice_Ride_2016_Station_Locations.csv")

```

Read in data on MetroTransit
```{r, cache=TRUE, message = FALSE, warning = FALSE}
metroStops<-read_csv("bus-stops-Oct-2017.csv")
metroRidership<-read_csv("ridership-route-day-Jan2014-Oct2017.csv")

```


Visualize where the bus stops and bike stations are in the twin cities
```{r, cache=TRUE,message = FALSE, warning = FALSE}
twinCitiesMap<-get_map(location="Minneapolis", zoom= 10)

ggmap(twinCitiesMap)+
   geom_point(data=metroStops, aes(x=site_longitude, y=site_latitude))



ggmap(twinCitiesMap)+
   geom_point(data=niceRideLocations, aes(x=Longitude, y=Latitude))


```


#Weather analysis

Read in data on weather for 2016
```{r, cache=TRUE}

#source: http://www.dnr.state.mn.us/climate/twin_cities/listings.html
#Had to copy and paste data from page into .txt file and read that in
MinneapolisWeather<-read_csv(file="Minneapolis Weather.txt", col_names = c("date", "maxTemp","minTemp","precip","snow","snowDepth"))
```


Clean up data
```{r}
MinneapolisWeather<-MinneapolisWeather%>%
   mutate(precip=as.double(replace(precip,precip=="T",0)))%>%
   mutate(snow=as.double(replace(snow,snow=="T",0)))%>%
   mutate(snowDepth=as.double(replace(snowDepth,snowDepth=="T",0)))

```

Show daily ridership from 2014-2017 overall
```{r}

#change date variable type to Date object
metroRidership<-metroRidership%>%
   mutate(dtDate=as.Date(dtDate))

#Calculate total number of riders for a given day, while still keeping data on what type of riding day it was (Holiday, weekday, etc.)
metroRidershipTotal<-metroRidership%>%
   group_by(dtDate)%>%
   summarize(totalRiders=sum(Total_Riders))%>%
   left_join(metroRidership, by="dtDate")%>%
   select(dtDate, totalRiders, Schedule)

#Show graph of total ridership for each day
ggplot(metroRidershipTotal, aes(x=dtDate, y=totalRiders, color=Schedule))+
   geom_point()

```


Show how divergence from daily mean effects ridership (only work with 2016 because thats time frame for nice ride data)
```{r}

#Add weather data to info on bus ridership
WeatherRiders<-metroRidership%>%
   left_join(MinneapolisWeather, by=c("dtDate"="date"))%>%
   filter(dtDate>="2016-01-01"&dtDate<="2016-12-31")#fix date object type


#Calculate the average temperature of a day over the seven years of weather data
#Should replace this with climate normals from NOAA
AvgDayTemp<-MinneapolisWeather%>%
   mutate(day=yday(date))%>%
   group_by(day)%>%
   summarize(dayTemp=sum((maxTemp+minTemp)/2)/n())

#Calculate the difference between the average temperature of a day and the mean average temperature of that day over the seven years of weather data 
WeatherRiders<-WeatherRiders%>%
   mutate(day=yday(dtDate))%>%
   left_join(AvgDayTemp, by="day")%>%
   mutate(tempDiff=(maxTemp+minTemp)/2-dayTemp)

WeatherRidersTotal<-WeatherRiders%>%
   mutate(day=as.factor(day), Route=as.factor(Route))%>%
   group_by(dtDate, tempDiff, dayTemp)%>%
   summarize(Total_Riders=sum(Total_Riders))


```


Visualization of how ridership is dependent on temperature of a day and how it depends on divergence from historical temperature on that day
```{r}

#Group together all the riders on the different routes while keeping temp data

ggplot(WeatherRidersTotal, aes(x=dayTemp, y=Total_Riders))+
   geom_smooth()
```

We see that ridership does correlate wtih absolute temperature peaking at around 50 degrees F
```{r}
ggplot(WeatherRidersTotal, aes(x=tempDiff, y=Total_Riders))+
   geom_smooth()+
   geom_point(alpha=.5)

```

```{r}
#How ridership varies based on the Max Temperature for any given day
ggplot(WeatherRiders, aes(x=maxTemp, y=Total_Riders))+
  geom_smooth()

#How ridership varies based on the Min Temperature for any given day
ggplot(WeatherRiders, aes(x=minTemp, y=Total_Riders))+
  geom_smooth()

#Visualize the relationship between the ridership patterns and the min and max temperatures of any given day (is a bit weird/unclear at first to interpret, so we might not wanna use this)
#Overall ridership seems to be higher for the min temperatures because that means it was likely a generally warmer day than when you are just looking at the max temp (because the temp could have dropped so much lower than that max temp during the day)
ggplot(WeatherRiders, aes(x=minTemp, y=Total_Riders))+
  geom_smooth()+
  geom_smooth(aes(x=maxTemp, y=Total_Riders, color="red"))+
  labs(x="Temperature")
```

Ridership drops off significantly on days where weather is significantly colder than norms, and is approximately constant for temps above norms


##Doing same weather analysis with ridership
First get ridership of bikes by day

```{r}

#clean up date variable in data and total riders on that date
niceRideDay<-niceRide2016%>%
   mutate(startDate=niceRide2016$`Start date`)%>%
   select("startDate")%>%
   mutate(startDate=as.Date(startDate, "%m/%d/%Y"))%>%
   group_by(startDate)%>%
   summarize(numRiders=n())%>%
   filter(!is.na(startDate))


#Add weather data to NiceRide data, including divergence from normals
niceRideDay<-niceRideDay%>%
   left_join(MinneapolisWeather, by=c("startDate"="date"))%>%
   mutate(day=yday(startDate))%>%
   left_join(AvgDayTemp,by="day")%>% 
   mutate(tempDiff=(maxTemp+minTemp)/2-dayTemp)


```

Look at how NiceRide ridership correlates to temperature
```{R}
ggplot(niceRideDay, aes(x=dayTemp, y=numRiders))+
   geom_smooth()
   
```

Unsuprisingly, more people ride bikes in warmer weather

Now look at how ridership correlates to departure from climate normals
```{R}
ggplot(niceRideDay, aes(x=tempDiff, y=numRiders))+
   geom_smooth()+
   geom_point(alpha=.5)

```
People ride much less on days with much lower temperatures than normals, and more on days higher than normals

```{r}

niceRideLatLongs<-matrix(c(niceRideLocations$Longitude, niceRideLocations$Latitude),ncol=2)
busLatLongs<-matrix(c(metroStops$site_longitude, metroStops$site_latitude),ncol=2)

minDist<-apply(distm(x=niceRideLatLongs, y=busLatLongs, fun = distHaversine),1, min)

niceRideLocationsDist<-niceRideLocations%>%
   mutate(minDist=minDist)

```


```{r}
niceRide2016dist<-niceRide2016%>%
   left_join(niceRideLocationsDist, by=c("Start station"="Station"))%>%
   group_by(Latitude, Longitude, minDist)%>%
   filter(!is.na(Latitude))


ggplot(niceRide2016dist, aes(x=minDist))+
   geom_density(fill="wheat")
```

```{r}
niceRide2016WDay<- niceRide2016 %>%
   mutate(`Start date`=as.Date(`Start date`, "%m/%d/%Y"))%>%
   mutate(DayOfWeek=wday(`Start date`))%>%
   filter(!is.na(`Start date`))%>%
   filter(`Account type`!= "Inconnu")


ggplot(niceRide2016WDay, aes(x= DayOfWeek)) + geom_bar(position = "dodge",aes(fill=`Account type`), alpha= 0.3)
```

```{r,message = FALSE, warning = FALSE}

niceRideProximityRiders<-niceRide2016%>%
  left_join(niceRideLocationsDist, by=c("Start station"="Station"))%>%
  group_by(`Start station`, Latitude, Longitude, `Account type`)%>%
  summarize(total=n())%>%
  arrange(desc(total))%>%
  filter(!is.na(Latitude))


niceRideProximityRiders2<-niceRide2016%>%
  left_join(niceRideLocationsDist, by=c("Start station"="Station"))%>%
  group_by(`Start station`, Latitude, Longitude)%>%
  summarize(total=n())%>%
  arrange(desc(total))%>%
  filter(!is.na(Latitude))

twinCitiesMap<-get_map(location="Minneapolis", zoom= 13)

ggmap(twinCitiesMap)+
   geom_point(data=niceRideProximityRiders2, aes(x=Longitude, y=Latitude, size=total, alpha=.5))

#Visualization of niceride stations in the twin cities- the bigger the circle, the more people used the station on the day 11/6/16.
```

```{r}
niceRide2016Dist<-niceRide2016%>%
   left_join(niceRideLocationsDist, by=c("Start station"="Station"))

ggplot(niceRide2016Dist, aes(x=minDist))+
   geom_density(fill="wheat")

#Visualize how the usage of the niceride bike stations based on their proximity to a bus station and if they are a casual user or member
```





```{r}

niceRideLatLongs<-matrix(c(niceRideLocations$Longitude, niceRideLocations$Latitude),ncol=2)
busLatLongs<-matrix(c(metroStops$site_longitude, metroStops$site_latitude),ncol=2)



distanceMatrix<-distm(x=niceRideLatLongs, y=busLatLongs, fun = distHaversine)

distanceDF<-as.data.frame(t(distanceMatrix))


colnames(distanceDF)<-niceRideLocations$Terminal


closeDF<-distanceDF%>%
   mutate(minDist=apply(distanceDF,1,min))%>%
   mutate(site_id=as.integer(metroStops$site_id))%>%
   filter(minDist<200)

```







```{r}

set.seed(1)
gatheredDist<-closeDF%>%
   gather(key=niceRideTerminal, value=distance, 1:202)%>%
   arrange(site_id)%>%
   select(-minDist)
   

siteSample<-sample(gatheredDist$site_id, size=3)


niceRide2016Clean<-niceRide2016%>%
   mutate(`Start date`=hour(mdy_hm(`Start date`))+minute(mdy_hm(`Start date`))/60)



niceRide2016Work<-niceRide2016Clean%>%
   #filter(`Account type`=="Member")%>%
   mutate(wday=wday(`Start date`))%>%
   filter(wday!=6|wday!=7)%>%
   filter(`Start date`<=20&`Start date`>=15)%>%
   select(-wday)



gatheredSampledClose<-gatheredDist%>%
   filter(site_id%in%siteSample)%>%
   filter(distance<500)


niceRideSampledClose<-niceRide2016Work%>%
   filter(`Start station number`%in% gatheredSampledClose$niceRideTerminal)%>%
   left_join(gatheredSampledClose, by=c("Start station number"="niceRideTerminal"))%>%
   mutate(site_id=as.factor(site_id))

```


```{r}
ggplot()+
   geom_density(data=niceRide2016Work, aes(x=`Start date`, ..density..), fill= "black", alpha=.5)+
   geom_freqpoly(data=niceRideSampledClose, binwidth=1/12 , aes(x=`Start date`, ..density..),color="black", boundary=15)+
   geom_freqpoly(data=niceRideSampledClose, binwidth=1/6 , aes(x=`Start date`, ..density.., color=`Start station number`), position = "identity", boundary=15, alpha=.3)+
   facet_grid(site_id~.)+
   labs(x="Start Time")

```
```{r, cache=TRUE}
metroStops<-read_csv("https://raw.githubusercontent.com/grahamammal/Data-Science-Final/master/bus-stops-Oct-2017.csv")
metroRidership<-read_csv("https://raw.githubusercontent.com/grahamammal/Data-Science-Final/master/ridership-route-day-Jan2014-Oct2017.csv")

metroRidership2016<-metroRidership%>%
  mutate(dtDate=as.Date(dtDate))
  

MinneapolisWeather
metroRidership2016

#Add weather data to info on bus ridership
WeatherRiders<-metroRidership2016%>%
   left_join(MinneapolisWeather, by=c("dtDate"="date"))%>%
   filter(dtDate>="2016-01-01"&dtDate<="2016-12-31") #fix date object type

ggplot(WeatherRiders, aes(x=snowDepth, y=Total_Riders))+
  geom_col(color="blue4")
#shows how snow depth on a given day affects the ridership of metro transit
```
